Hire Ed: Job Market Dynamics for Tenure-Track Faculty Positions in Archaeology

### 2024-10-22 analysis/paper/article.qmd
writing the discussion

@@

Panel B of @fig-show-map-of-hiring-institution-and-cc shows the geographic distribution of the hiring institutions. California posted almost twice as many job ads as the next most active states. After California, the states that posted the most ads during 2013-2023 include New York, Texas, and Pennsylvania, and Florida. These top five states also correspond to the top five most populous US states, indicating that rates of hiring is approximately proportional to population density. Similarly, the lowest counts of job ads were observed in states with the lowest populations: North Dakota, South Dakota, Alaska, and Nebraska. No institutions in Montana posted a job ad during this period.

@@

In the years 2019-2020 and 2020-2021 there are striking changes in the popularity of topics in job ads. Indigenous and historical archaeology became the most popular topic at this time, rising from being one of the least popular topics from 2012-2017. Conversely, archaeological science, which was popular during 2012-2017, was rarely mentioned in job ads during 2019-2021. Ancient Europe and the Mediterranean, which was consistently one of the least frequently mentioned topics, virtually disappeared from job ads during 2019-2021. Mentions  of complex societies in ads consistently decrease over the study period, while mentions of public archaeology consistently increase. The topic of Environmental Archaeology remains consistently high over time.

@@

@fig-requirements-by-position

@@

>>>>

<!--

cite these

- https://www.anthropology-news.org/articles/hello-to-tristes-tropes/

@@

Panel B of @fig-show-map-of-hiring-institution-and-cc shows the geographic distribution of the hiring institutions. California posted almost twice as many job ads as the next most active states. After California, the states that posted the most ads during 2013-2023 include New York, Texas, and Pennsylvania, and Florida. These top five states also correspond to the top five most populous US states, indicating that rates of hiring is approximately proportional to population density. These top five states for job ads are also the top five states with highest number of degrees awarded in Anthropology (U.S. Department of Education nd). Similarly, the lowest counts of job ads were observed in states with the lowest populations: North Dakota, South Dakota, Alaska, and Nebraska. No institutions in Montana posted a job ad during this period.

@@

In the years 2019-2020 and 2020-2021 there are striking changes in the popularity of topics in job ads. Indigenous and historical archaeology (which includes archaeology of enslaved people) became the most popular topic at this time, rising from being one of the least popular topics from 2012-2017. Conversely, archaeological science, which was popular during 2012-2017, was rarely mentioned in job ads during 2019-2021. Ancient Europe and the Mediterranean, which was consistently one of the least frequently mentioned topics, virtually disappeared from job ads during 2019-2021. Mentions  of complex societies in ads consistently decrease over the study period, while mentions of public archaeology consistently increase. The topic of Environmental Archaeology remains consistently high over time.

@@

Instructions to applicants vary by the rank of the position advertised [@fig-requirements-by-position]. Assistant professor positions tend to have the most complex instructions with many different types of documents required. Conversely, full professor positions have minimal requirements. The documents that are most frequently requested across the three ranks are cover letter and CV.

@@

Our results reveal some of these shifts, providing insights into the dynamics of topical, methods, and geographic foci over time. While these dynamics are deeply felt by people actively seeking faculty jobs, we believe this is the first time they have been quantified at such a large scale. Trends in job ads reflect broader shifts in intellectual and practical priorities about archaeology, undergraduate education, and the process of hiring professors. The demand for archaeology faculty, indicated by the total number of tenure-track jobs, maybe explained by a variety of processes. The downward trend in tenure-track positions during 2013-2019 related to decline in undergraduate enrollment in anthropology fields since 2013 [@crambChangingProfileTenureTrack2022a]. The big dip during 2020-2021 is explained by the hiring freezes at many institutions resulting from the COVID-19 pandemic, which caused extreme disruption and uncertainty as universities adapted to online instruction and assessment in an effort to minimize the spread of the virus. 
In addition to a dip in job ads, our results show that 2019-2021 was a major inflection point in the popularity of topics and regions in job ads. Indigenous and historical archaeology and the Americas strongly increase in popularity at this time, while archaeological science, complex societies, the Mediterranean and Near East and show declines. Similarly, the number of job ads with a geographic focus on the Americas and Africa peaks during this time. These shifts in the topics and geographic foci of job ads were likely been influenced by broader cultural movements such as Black Lives Matter, protests about racial injustice, and efforts to amplify Indigenous voices. COVID-19 negatively impacted Black, American Indian and Hispanic communities with significantly higher infection and morbidity rates, drawing attention to racial and socio-economic inequality in the US. The Black Lives Matter movement, dating back to 2013, intersected profoundly with the pandemic with the murder of George Floyd by a police officer in May 2020, three months after the World Health Organization declared COVID-19 to be a pandemic. Mass protests objecting to Floyd's murder generated widespread concern about racial inequities and stimulated a broad interest in addressing systemic racial injustice. Our data suggest that US universities participated in this movement by adjusting their hiring plans to prioritize recruiting archaeologists working on topics relevant to Black and Indigenous communities. Many universities may have hoped to hire Black and Indigenous archaeologists as part of their effort to tackle systemic racism. However as it is illegal in the US for universities use race or ethnicity as a primary factor in hiring (due to the 1964 Civil Rights Act), universities tailored the content of the job ads to focus on topics where Black and Indigenous researchers are most numerous, ensuring that these researchers would be well-represented in the pool of applicants.

This striking change in topics and geographic foci of 2019-2021 occurred against a backdrop of several longer-term trends. We found a long-term decline in topics such as the archaeology of complex societies and archaeological science, Mesoamerica and South America as a geographic focus, and methods relating to landscape archaeology. These trends are harder to explain as we cannot link their start to a discrete moment like COVID-19. We might speculate that a growing preference for archaeological approaches that privilege agency-driven, relational perspectives in combination with human subjectivity and cultural framing, is one enduring legacy of debates in the 1980s and 90s about processualism versus post-processualism. This might explain why archaeological science is showing a decline, as demand for methods for analyzing artefact materiality, ontology, and power displace physical methods for compositional analysis. A decline in interest in the archaeology of complex societies may reflect several themes that intersect with broader social changes, such as growing interest in Indigenous and non-state actors in the past, increased interest in climate change, environmental sustainability, and resilience in the past may have shifted attention away from the study of monumental architecture, elite societies, political hierarchies and state systems. Other relevant factors may include increasing difficulty of obtaining research funding to support archaeological science research, such as laboratory facilities and instrumentation, and an increasing emphasis on more job-oriented degrees that prepare students for working in cultural heritage management rather than advanced research.

Our data on the requirements for applicants shows that application complexity has gradually increased over time. This trend is especially pronounced for Assistant Professor positions which make more demands on applicants than Associate and Full Professor positions. We saw a growing demand for research and teaching statements. Demand for diversity statements shows a unique trajectory, peaking in 2020-2021 and declining into the present. This may relate to the intersecting concerns emerging during the COVID-19 pandemic, which have declined in job ads as the pandemic moves into the past. The decline in recent years for names of recommenders in the initial application may be a response to criticisms of the burden on the applicant of preparing numerous complex job applications [e.g. @dennisWorstAnthroJob2022]. In recognition of this burden, not only on applicants but also on colleagues writing letters of recommendation over and over to support applicants, many hiring committees now follow the recommendations of @dennisWorstAnthroJob2022 in only requesting names and letters of recommendation at later stages of the hiring process, if at all.

A key limitation of our research is that it does not include an analysis of the academic profiles of those who were eventually hired from the job ads. Our results reveal collective aspirations for the future of the field, but do not show how these worked out in the topical and geographic foci of the people who were eventually hired.
